Algebraic Graphs with Class

### 2021-05-31 paper/paper.tex
Finish paper

\section{Introduction}\label{sec:intro}
Graphs are a fundamental structure studied in depth by both mathematicians and
computer scientists alike. One very common definition states that a (directed)
graph is a tuple $G = (V,E)$, where $V$ is a set of vertices and $E \subseteq V
\times V$ is the set of edges. While this is a perfectly valid and natural
mathematical definition it is not necessarily suitable for implementation. This
can be illustrated by trying to directly translate this to Haskell:
\begin{minted}{haskell}
  data G a = G { vertices :: Set a, edges :: Set (a,a)}
\end{minted}
The value \hs{G [1,2,3] [(1,2),(2,3)]} then represents the graph $G =
(\{1,2,3\}, \{(1,2),(2,3)\})$, however \hs{G [1,2] [(2,3)]} does not
represent a consistent graph, as the edge refers to a non-existent node.
The state-of-the-art \texttt{containers} library implements graphs with
adjacency lists employing immutable arrays~\cite{king1995dfs}. The
consistency condition $E \subseteq V \times V$ is not checked statically though,
which can lead to runtime errors. \texttt{fgl}, another popular Haskell graph
library uses inductive graphs~\cite{erwig2001inductive}, also exhibiting partial
functions and potential for runtime errors due to the violation of consistency.

@@

Alongside proper definitions for these primitives we will see
in~\autoref{sec:locale} that this is indeed a sound and complete graph
representation. We will also cover the algebraic structure (\ref{sec:algebra})
exhibited by these graphs and elegant graph transformations (\ref{sec:trafo})
based on a type class for the core. In~\autoref{sec:verification} we explore how
we can exploit the algebraic structure in verification using Isabelle/HOL.

@@

The construction primitives
then are defined as follows. The \textit{empty} graph, denoted by $\eps$, is
simply the tuple $(\emptyset, \emptyset)$ which is clearly consistent, hence
$\eps \in \mathcal{G}$. Graphs with a single \textit{vertex} $v \in
\mathbb{V}$ and no edges, i.e.\ $(\{v\}, \emptyset)$, denoted by just $v$, are also
trivially consistent, hence $v \in \mathcal{G}$. The binary operations
\textit{overlay} and \textit{connect}, denoted by $\overlay$ and $\connect$
respectively, allow constructing larger graphs. Their definitions are
\begin{align*}
  (V_1, E_1) \overlay (V_2, E_2) &\coloneqq (V_1 \cup V_2, E_1 \cup E_2)\\
  (V_1, E_1) \connect (V_2, E_2) &\coloneqq (V_1 \cup V_2, E_1 \cup E_2 \cup V_1 \times V_2)
\end{align*}
So overlay is just the union of two graphs, while connect also adds an edge from
each vertex of its first argument to each vertex of its second argument. It is
straightforward to see that if $G_1 \in \mathcal{G}$ and $G_2 \in \mathcal{G}$
are consistent, then so are $G_1 \overlay G_2 \in \mathcal{G}$ and $G_1 \connect
G_2 \in \mathcal{G}$. In combination with the consistency of $\eps$ and of single vertex
graphs this already proves that algebraic graphs are a consistent
representation for graphs, i.e.\ using the construction primitives only yields
consistent graphs. The completeness of this representation, i.e.\ for
each $G \in \mathcal{G}$ there is an algebraic graph expression representing it,
is shown shortly (\ref{sec:algebra}). Both of these facts were also formally proven in
Isabelle/HOL (see~\autoref{sec:verification}).

@@

\subsection{Algebraic Structure}\label{sec:algebra}
Before turning to more advanced graph constructions we will observe the
algebraic structure of the primitives. It turns out that overlay and connect
almost form a semiring over graphs:
\begin{itemize}
\item $(\mathcal{G}, \overlay, \eps)$ is an idempotent, commutative monoid
\item $(\mathcal{G}, \connect, \eps)$ is a monoid
\item $\connect$ distributes over $\overlay$
\end{itemize}
For a proper semiring an annihilating zero for $\connect$ (i.e.\
$x \connect 0 = 0 = 0 \connect x$) is missing. It is also unusual that the two operations
share their identity $\eps$.

In addition there is the following
\textit{decomposition law}, as illustrated in~\autoref{fig:axioms}(b).
\[
  x \connect y \connect z = x \connect y \overlay x \connect z \overlay y \connect z
\]
\begin{figure*}
  \begin{subfigure}[b]{0.4\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-distributivity.pdf}}
    \caption{Distributivity: $1 \connect (2 \overlay 3) = 1 \connect 2 \overlay
      1 \connect 3$ }
  \end{subfigure}
  \hspace{12mm}
  \begin{subfigure}[b]{0.5\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-decomposition.pdf}}
    \caption{Decomposition: $1 \connect 2 \connect 3 = 1 \connect 2 \overlay
      {1 \connect 3} \overlay {2 \connect 3}$}
  \end{subfigure}
  \vspace{-1mm}
  \caption{Two axioms of the algebra of graphs. Illustrations from~\cite{mokhov2017algebraic}.\label{fig:axioms}}
\end{figure*}
With decomposition we can prove by equational reasoning that $\overlay$ is
idempotent and that $\eps$ is indeed its identity.
According to Mokhov, the remaining axioms are a minimal set of axioms
\emph{characterizing} algebraic graphs~\cite{mokhov2017algebraic}. From the
original paper it does not become clear what he means by characterization:
Letting $\mathcal{G} = \{1\}$, $\eps = 1$, $1 \overlay 1 = 1$ and
$1\connect 1 = 1$ gives a structure abiding all axioms. This definition of
$\mathcal{G}$ is clearly not (isomorphic to) what we call directed graphs for
any universe of vertices.
We don't know which additional axioms are required to obtain this isomorphism.
However, the statement $\eps \neq vertex\ u$ for any $u \in \mathbb{V}$ does not
follow from the given axioms. Hence, adding an axiom scheme $\forall u \in
\mathbb{V}: \eps \neq vertex\ u$ seems sensible. Additionally $vertex$ probably
needs to be injective (i.e.\ $vertex\ u = vertex\ v \implies u = v$), however, we don't know if this
is sufficient.

One can, however, easily verify that the definitions of the primitives given
in~\autoref{sec:core} satisfy these axioms. At this point let us consider the
completeness of algebraic graphs. For any graph $G=(V,E)$ the following function
builds it using the four construction primitives:
\begin{minted}{haskell}
  graph :: Graph g => [Vertex g] -> [(Vertex g, Vertex g)] -> g
  graph vs es = overlay (vertices vs) (edges es)
\end{minted}
The \hs{edges} function generalizes \hs{edge} to a list of edges:
\begin{minted}{haskell}
  edges :: Graph g => [(Vertex g, Vertex g)] -> g
  edges = foldr overlay empty . map (uncurry edge)
\end{minted}
They are a direct translation of the fact, that any algebraic graph expression
$g$ for a graph $G=(V,E) \in \mathcal{G}$ can be rewritten in the 
\textit{canonical form}:
\[
  g = \left( \bigoplus_{v \in V} v \right) \oplus \left( \bigoplus_{(u,v) \in E}
  u \connect v \right)
\]

@@

So far we only considered directed graphs. Undirected graphs can be
obtained from the very same construction primitives as directed graphs. It suffices
to modify the underlying axioms. In the case of undirected graphs this can be
achieved by making connect commutative (then denoted by $\leftrightarrow$),
i.e.\ adding the axiom $x \leftrightarrow y = y \leftrightarrow x$. By
introducing this axiom we can prove some of the other axioms of directed graphs.
Hence, according to Mokhov~\cite{mokhov2017algebraic}, undirected graphs are
characterized by the following minimal set of axioms:
\begin{itemize}
\item $(\mathcal{G}, \oplus)$ is a commutative semigroup
\item $\leftrightarrow$ is commutative and has $\eps$ as the identity
\item left distributivity: $x \leftrightarrow (y \oplus z) = x \leftrightarrow
  y \oplus x \leftrightarrow z$
\item left decomposition: $(x \leftrightarrow y) \leftrightarrow z = x
  \leftrightarrow y \oplus x \leftrightarrow z \oplus y \leftrightarrow z$
\end{itemize}
By choosing different additional axioms, other classes of graphs can be
characterized. Some examples are reflexive graphs (i.e.\ each vertex has a
self-loop), transitive graphs (as used in dependency graphs), and the combination
thereof; preorders. Combining undirected graphs and reflexive graphs yields
equivalence relations. It is even possible to define hypergraphs of arbitrary
(but fixed) rank by replacing the decomposition law with another suitable
axiom~\cite{mokhov2017algebraic}. In all of these the same caveats regarding the
term \emph{characterization} from directed graphs apply.

@@

  instance Ord a => Graph (Relation a) where
    type Vertex (Relation a) = a
    empty       = R Set.empty Set.empty
    vertex  x   = R (singleton x) Set.empty
    overlay x y = R (domain x `union` domain y) (relation x `union` relation y)
    connect x y = R (domain x `union` domain y) (relation x `union` relation y `union`
      fromAscList [ (a, b) | a <- elems (domain x), b <- elems (domain y) ])
\end{minted}
Mokhov also defines a \hs{Num} instance for \hs{Relation} which enables the use
of \hs{+} and \hs{*} as shortcuts for \hs{overlay} resp.\ \hs{connect}.

@@

  instance Graph g => Graph (Transpose g) where
    type Vertex (Transpose g) = Vertex g
    empty       = T empty
    vertex      = T . vertex
    overlay x y = T $ overlay (transpose x) (transpose y)
    connect x y = T $ connect (transpose y) (transpose x)
\end{minted}
Although for certain representations it is possible to get an $\mathcal{O}(1)$
transpose operation (e.g.\ by setting a transposed flag), for many typical
representations the whole graph has to be traversed, resulting in
$\mathcal{O}(\abs{V} + \abs{E})$ time.

More sophisticated transformations can be implemented using the following
\hs{Functor}-like \hs{newtype} wrapper. The function \hs{gmap} takes a function
\hs{a -> b} and a graph with vertices of type \hs{a} and produces a graph with
vertices of type \hs{b}.
\begin{minted}{haskell}
  newtype GraphFunctor a = F { gfor :: forall g. Graph g => (a -> Vertex g) -> g }

@@

  gmap :: Graph g => (a -> Vertex g) -> GraphFunctor a -> g
  gmap = flip gfor  
\end{minted}
Interestingly this can be used to merge vertices fulfilling some predicate
\hs{p} by mapping them to a single vertex:
\begin{minted}{haskell}
  mergeVertices :: Graph g => (Vertex g -> Bool)
    -> Vertex g -> GraphFunctor (Vertex g) -> g
  mergeVertices p v = gmap $ \u -> if p u then v else u
\end{minted}
In the original paper \hs{gmap} is then used further to construct rather
sophisticated graphs fully polymorphically~\cite{mokhov2017algebraic}, i.e.\
independent from any concrete graph representation.

@@

To experiment further we first formalized the material from the paper in
Isabelle/HOL~\cite{isabelle}.
This formalization\footnote{available at
  \texttt{\href{https://github.com/cmadlener/algebraic_graphs}{https://github.com/cmadlener/algebraic\_graphs}}}
the definition of several \textit{locale}s. For the purposes of this paper a
locale be thought of as a type class which can be equipped with axioms. The
following locale is the basis for our formalization.

@@

As the statements in the original paper were of very limited scope, we 
define and reason about more \textit{advanced} graph theoretic concepts: We
introduce (vertex) walks and the notion of reachability on algebraic graphs. A
challenge posed by the approach is that a graph is simply a value of some
abstract type \Term{'g}. Thus, we characterize the existence of a vertex or an
edge in a given graph by the subgraph relation (denoted by $\sqsubseteq$ in the
formalization):

\vspace{3mm}
\term{terms/has_elem}
\vspace{3mm}\newline
Equipped with these definitions we can then go on to inductively define when a
list of vertices is a walk in a given graph:

\vspace{3mm}
\term{terms/vwalk_def}
\vspace{3mm}\newline
We define the empty list as well as the list containing only a single vertex of
the graph as walks (\Term{vwalk\_Nil} and \Term{vwalk\_single}). Additionally if
a list with head \Term{v} (note that Haskell's \hs{(:)} is denoted by \Term{(\#)} in
Isabelle) is a walk and $(u,v) \in E$, then prepending \Term{u} also yields a
walk. Now we can prove various statements about walks (e.g.\ appending walks,
splitting walks, etc.). Interestingly, many of these statements follow directly
from the inductive definition of walks, i.e.\ we don't require any of the
algebraic graph axioms to prove them. We only need them when a proof relies on
the fact that $(u,v) \in E$ implies $u \in V$ as well as $v \in V$ (in other
words, the consistency of a graph).

@@

\vspace{3mm}
\term{terms/edge_set_def}
\vspace{3mm}\newline
Reachability then can be defined via the reflexive, transitive closure of the
edge set. The second approach is to say that $v$ is reachable from $u$ if and
only if there is a walk from $u$ to $v$. These are, however, only proof of
concepts as only transitivity of reachability has been shown for both definitions.

Nonetheless, first impressions from these experiments suggest that it is
feasible to base more sophisticated formalizations of graph theory on algebraic
graphs. We did not experiment with adding additional axioms as described
in~\autoref{sec:algebra}. We instantiate the locale with the unit type, where
all primitives simply map to the unit value. This formalizes that the current
set of axioms is insufficient.

@@

\section{Deep Embedding}\label{sec:deep}
Up until this point we were mainly concerned with exploiting the algebraic
structure of graphs for polymorphic graph constructions and transformations, and
for formal verification. However, the deep embedding of the
construction primitives in a datatype is interesting in itself. Recall the
definition from the introduction:
\begin{minted}{haskell}
  data Graph a = Empty
               | Vertex a
               | Overlay (Graph a) (Graph a)
               | Connect (Graph a) (Graph a)
\end{minted}
An instance for the type class can be defined easily, however, we cannot use the
derived \hs{Eq} instance for the datatype, as it would violate the graph axioms:
\hs{Overlay Empty Empty} and \hs{Empty} both represent $\eps$ but are not deemed
equal.
One solution is to fold the deep embedding into a concrete graph representation to
obtain a law-abiding \hs{Eq} instance:
\begin{minted}{haskell}
  fold :: Graph g => Graph (Vertex g) -> g
  fold Empty         = empty
  fold (Vertex  x  ) = vertex x
  fold (Overlay x y) = overlay (fold x) (fold y)
  fold (Connect x y) = connect (fold x) (fold y)

@@

The quotient type also allows us to formally prove that it satisfies the axioms,
by instantiating the locale corresponding to the core type class. It also paves
the way for further formalization efforts on the deep embedding. However, this
also requires more work to become truly interesting, i.e.\ if algorithms that
work directly on the deep embedding or other advanced usages of it can be found.

@@

Algebraic graphs as presented have been successfully used in practice, e.g.\ for
asynchronous circuit design~\cite{beaumont2017high} and acceleration of graph
processing using FPGAs~\cite{mokhov2019language}. In these examples
the available transformations are sufficient. An important direction for future
research is going beyond those (relatively simple) transformations. That means
to find and formulate algorithms either inside the core type class or directly
on the deep embedding. Berghammer et. al~\cite{berghammer2020relational}, as well
as, Guttmann and Robinson-O'Brien~\cite{Relational_Minimum_Spanning_Trees-AFP}
successfully employ algebraic approaches to formally verify graph algorithms.
However, they are using an algebra of relations as opposed to the one here
defined on graph construction primitives. Dolan~\cite{dolan2013fun} uses an
algebra on adjacency matrices (specifically a semiring) to find paths in graphs.
Mokhov himself postulates that the algebraic approach allows to formulate graph
algorithms as systems of equations with unknowns. This may also be an approach
for finding novel graph algorithms.

>>>>

\section{Introduction}\label{sec:intro}
Graphs are a fundamental structure studied in depth by both mathematicians and
computer scientists alike. One very common definition states that a (directed)
graph is a tuple $G = (V,E)$, where $V$ is a set of vertices and $E \subseteq V
\times V$ is the set of edges. While this is a perfectly valid and natural
mathematical definition it is not necessarily suitable for implementation. This
can be illustrated by trying to directly translate this to Haskell:
\begin{minted}{haskell}
data G a = G { vertices :: Set a, edges :: Set (a,a)}
\end{minted}
The value \hs{G [1,2,3] [(1,2),(2,3)]} then represents the graph $G =
(\{1,2,3\}, \{(1,2),(2,3)\})$, however \hs{G [1,2] [(2,3)]} does not
represent a consistent graph, as the edge refers to a non-existent node.
The state-of-the-art \texttt{containers} library implements graphs with
adjacency lists employing immutable arrays~\cite{king1995dfs}. The
\emph{consistency condition} $E \subseteq V \times V$ is not checked statically,
which can lead to runtime errors. \texttt{fgl}, another popular Haskell graph
library uses inductive graphs~\cite{erwig2001inductive}, also exhibiting partial
functions and potential for runtime errors due to the violation of consistency.

@@

Alongside proper definitions for these primitives we will see
in~\autoref{sec:locale} that this is indeed a sound and complete graph
representation. We will also cover the algebraic structure (\ref{sec:algebra})
exhibited by these primitives and elegant graph transformations (\ref{sec:trafo})
based on a type class for the core. In~\autoref{sec:verification} we explore how
we can exploit the algebraic structure in verification using Isabelle/HOL.

@@

The construction primitives
then are defined as follows. The \textit{empty} graph, denoted by $\eps$, is
the tuple $(\emptyset, \emptyset)$ which is clearly consistent, hence
$\eps \in \mathcal{G}$. Graphs with a single \textit{vertex} $v \in
\mathbb{V}$ and no edges, i.e.\ $(\{v\}, \emptyset)$, denoted by just $v$, are also
trivially consistent, hence $v \in \mathcal{G}$. The binary operations
\textit{overlay} and \textit{connect}, denoted by $\overlay$ and $\connect$
respectively, allow constructing larger graphs. Their definitions are
\begin{align*}
  (V_1, E_1) \overlay (V_2, E_2) &\coloneqq (V_1 \cup V_2, E_1 \cup E_2)\\
  (V_1, E_1) \connect (V_2, E_2) &\coloneqq (V_1 \cup V_2, E_1 \cup E_2 \cup V_1 \times V_2)
\end{align*}
Overlay is just the union of two graphs, while connect also adds an edge from
each vertex of its first argument to each vertex of its second argument. It is
straightforward to see that if $G_1 \in \mathcal{G}$ and $G_2 \in \mathcal{G}$
are consistent, then so are $G_1 \overlay G_2 \in \mathcal{G}$ and $G_1 \connect
G_2 \in \mathcal{G}$. In combination with the consistency of $\eps$ and of single vertex
graphs this already proves that algebraic graphs are a consistent
representation for graphs, i.e.\ the construction primitives only yield
consistent graphs. The completeness of this representation, i.e.\ for
each $G \in \mathcal{G}$ there is an algebraic graph expression representing it,
is shown shortly (\ref{sec:algebra}). Both of these facts were also formally proven in
Isabelle/HOL (see~\autoref{sec:verification}).

@@

\subsection{Algebraic Structure}\label{sec:algebra}
Before turning to more advanced graph constructions we will observe the
algebraic structure of the primitives. Overlay and connect form an algebra
abiding the following axioms:
\begin{itemize}
\item $\oplus$ is commutative and associative
\item $(\mathcal{G}, \connect, \eps)$ is a monoid
\item $\connect$ distributes over $\overlay$
\end{itemize}
In addition there is the following
\textit{decomposition law}, as illustrated in~\autoref{fig:axioms}(b).
\[
  x \connect y \connect z = x \connect y \overlay x \connect z \overlay y \connect z
\]
\begin{figure*}
  \begin{subfigure}[b]{0.4\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-distributivity.pdf}}
    \caption{Distributivity: $1 \connect (2 \overlay 3) = 1 \connect 2 \overlay
      1 \connect 3$ }
  \end{subfigure}
  \hspace{12mm}
  \begin{subfigure}[b]{0.5\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-decomposition.pdf}}
    \caption{Decomposition: $1 \connect 2 \connect 3 = 1 \connect 2 \overlay
      {1 \connect 3} \overlay {2 \connect 3}$}
  \end{subfigure}
  \vspace{-1mm}
  \caption{Two axioms of the algebra of graphs. Illustrations from~\cite{mokhov2017algebraic}.\label{fig:axioms}}
\end{figure*}
With decomposition we can prove by equational reasoning that $\overlay$ is
idempotent and that $\eps$ is its identity. Thus, overlay and connect
almost form a semiring over graphs:
\begin{itemize}
\item $(\mathcal{G}, \overlay, \eps)$ is an idempotent, commutative monoid
\item $(\mathcal{G}, \connect, \eps)$ is a monoid
\item $\connect$ distributes over $\overlay$
\end{itemize}
For a proper semiring the annihilating zero for $\connect$ (i.e.\
$x \connect \eps = \eps = \eps \connect x$) is missing. It is also unusual that the two operations
share their identity $\eps$.

According to Mokhov, the stated axioms are a minimal set of axioms
\emph{characterizing} algebraic graphs~\cite{mokhov2017algebraic}. From the
original paper it does not become clear what he means by characterization:
Letting $\mathcal{G} = \{1\}$, $\eps = 1$, $1 \overlay 1 = 1$ and
$1\connect 1 = 1$ gives a structure abiding all axioms. This definition of
$\mathcal{G}$ corresponds to graphs with an empty universe of vertices (as there
is only the empty graph). However, this definition is independent of the
underlying universe of vertices. So we can define the same structure
for a non-empty universe. In this scenario the primitives don't behave like what we
call (isomorphic to) graphs.
We don't know which additional axioms are required to obtain this isomorphism.
However, the statement $\eps \neq vertex\ u$ for any $u \in \mathbb{V}$ does not
follow from the given axioms. Hence, adding an axiom scheme $\forall u \in
\mathbb{V}: \eps \neq vertex\ u$ seems sensible. Additionally $vertex$ 
needs to be injective (i.e.\ $vertex\ u = vertex\ v \implies u = v$).

One can, however, easily verify that the definitions of the primitives given
in~\autoref{sec:core} satisfy these axioms. At this point let us consider the
completeness of algebraic graphs. For any graph $G=(V,E)$ the following function
builds it using the four construction primitives:
\begin{minted}{haskell}
graph :: Graph g => [Vertex g] -> [(Vertex g, Vertex g)] -> g
graph vs es = overlay (vertices vs) (edges es)
\end{minted}
The \hs{edges} function generalizes \hs{edge} to a list of edges:
\begin{minted}{haskell}
edges :: Graph g => [(Vertex g, Vertex g)] -> g
edges = foldr overlay empty . map (uncurry edge)
\end{minted}
They are a direct translation of the fact that any algebraic graph expression
$g$ for a graph $G=(V,E) \in \mathcal{G}$ can be rewritten in the 
\textit{canonical form}:
\[
  g = \left( \bigoplus_{v \in V} v \right) \oplus \left( \bigoplus_{(u,v) \in E}
    u \connect v \right)
\]

@@

So far we only considered directed graphs. Undirected graphs can be
obtained from the very same construction primitives as directed graphs. It suffices
to modify the underlying axioms. In the case of undirected graphs this can be
achieved by making connect commutative (then denoted by $\leftrightarrow$),
i.e.\ adding the axiom $x \leftrightarrow y = y \leftrightarrow x$. By
introducing this axiom we can prove some of the other axioms of directed graphs.
Hence, according to Mokhov~\cite{mokhov2017algebraic}, undirected graphs are
characterized by the following minimal set of axioms:
\begin{itemize}
\item $\oplus$ is commutative and associative
\item $\leftrightarrow$ is commutative and has $\eps$ as the identity
\item left distributivity: $x \leftrightarrow (y \oplus z) = x \leftrightarrow
  y \oplus x \leftrightarrow z$
\item left decomposition: $(x \leftrightarrow y) \leftrightarrow z = x
  \leftrightarrow y \oplus x \leftrightarrow z \oplus y \leftrightarrow z$
\end{itemize}
By choosing different additional axioms, other classes of graphs can be
characterized. Some examples are reflexive graphs (i.e.\ each vertex has a
self-loop), transitive graphs (as used in dependency graphs), and the combination
thereof; preorders. Combining undirected graphs and reflexive graphs yields
equivalence relations. It is even possible to define hypergraphs of arbitrary
(but fixed) rank by replacing the decomposition law with another suitable
axiom~\cite{mokhov2017algebraic}. In all of these the same caveats regarding the
term \emph{characterization} from directed graphs apply.

@@

instance Ord a => Graph (Relation a) where
  type Vertex (Relation a) = a
  empty       = R Set.empty Set.empty
  vertex  x   = R (singleton x) Set.empty
  overlay x y = R (domain x `union` domain y) (relation x `union` relation y)
  connect x y = R (domain x `union` domain y) (relation x `union` relation y `union`
    fromAscList [ (a, b) | a <- elems (domain x), b <- elems (domain y) ])
\end{minted}
Mokhov also defines a \hs{Num} instance for \hs{Relation} which enables the use
of \hs{+} and \hs{*} as shortcuts for \hs{overlay} and \hs{connect}.

@@

instance Graph g => Graph (Transpose g) where
  type Vertex (Transpose g) = Vertex g
  empty       = T empty
  vertex      = T . vertex
  overlay x y = T $ overlay (transpose x) (transpose y)
  connect x y = T $ connect (transpose y) (transpose x)
\end{minted}
$\mathcal{O}(1)$ transposition can be achieved by representations which allow to
use a flag to treat edges as transposed. For many typical representations (e.g.\
adjacency lists) the whole graph has to be traversed during transposition,
resulting in an $\mathcal{O}(\abs{V} + \abs{E})$ time operation.

More sophisticated transformations can be implemented using the following
\hs{Functor}-like \hs{newtype} wrapper. The function \hs{gmap} takes a function
of type \texttt{a -> b} and a graph with vertices of type \hs{a} and produces a graph with
vertices of type \hs{b}.
\begin{minted}{haskell}
newtype GraphFunctor a = F { gfor :: forall g. Graph g => (a -> Vertex g) -> g }

@@

gmap :: Graph g => (a -> Vertex g) -> GraphFunctor a -> g
gmap = flip gfor  
\end{minted}
Interestingly this can be used to contract vertices fulfilling some predicate
\hs{p} by mapping them to a single vertex:
\begin{minted}{haskell}
mergeVertices :: Graph g => (Vertex g -> Bool)
  -> Vertex g -> GraphFunctor (Vertex g) -> g
mergeVertices p v = gmap $ \u -> if p u then v else u
\end{minted}
In the original paper \hs{gmap} is then used further to construct rather
sophisticated graphs fully polymorphically~\cite{mokhov2017algebraic}, i.e.\
independent from any concrete graph representation.

@@

To experiment further we first formalized the material from the paper in
Isabelle/HOL~\cite{isabelle}.
This formalization\footnote{available at
  \texttt{\href{https://github.com/cmadlener/algebraic_graphs}{https://github.com/cmadlener/algebraic\_graphs}}}
includes several \textit{locale}s. For the purposes of this paper a
locale can be thought of as a type class which can be equipped with axioms. The
following locale is the basis for our formalization.

@@

As the statements in the original paper were of very limited scope, we 
define and reason about more \textit{advanced} graph theoretic concepts: We
introduce (vertex) walks and the notion of reachability on algebraic graphs. A
challenge posed by the approach is that a graph is simply a value of some
type \Term{'g}. Thus, we characterize the existence of a vertex or an
edge in a given graph by the subgraph relation.
A graph $g$ contains a vertex $u$ if and only if $u \subseteq g$ and an edge
$(u,v)$ if and only if $u \connect v \subseteq g$. The Isabelle definitions for
this look as follows:

\vspace{3mm}
\term{terms/has_elem}
\vspace{3mm}\newline
Where $\sqsubseteq$ is used instead of $\subseteq$ and \Term{edge u v} is
defined as \Term{connect (vertex u) (vertex v)}.
Equipped with these definitions we can then go on to inductively define when a
list of vertices is a walk in a given graph:

\vspace{3mm}
\term{terms/vwalk_def}
\vspace{3mm}\newline
We define the empty list as well as the list containing only a single vertex of
the graph as walks (\Term{vwalk\_Nil} and \Term{vwalk\_single}). Additionally if
a list with head \Term{v} (note that Haskell's \hs{(:)} is denoted by \Term{(\#)} in
Isabelle) is a walk and $(u,v) \in E$, then prepending \Term{u} also yields a
walk. Now we can prove various statements about walks (e.g.\ under which
conditions walks can be appended, how walks can be split, etc.). Interestingly,
many of these statements follow directly
from the inductive definition of walks, i.e.\ we don't require any of the
algebraic graph axioms to prove them. We only need them when a proof relies on
the fact that $(u,v) \in E$ implies $u \in V$ as well as $v \in V$ (in other
words, the consistency of a graph).

@@

\vspace{3mm}
\term{terms/edge_set_def}
\vspace{3mm}\newline
Reachability then can be defined via the reflexive, transitive closure of the
edge set, while also considering every vertex can reach itself. The second approach is to say that $v$ is reachable from $u$ if and
only if there is a walk from $u$ to $v$. These are, however, only proof of
concepts as only transitivity of reachability has been shown for both definitions.

Nonetheless, first impressions from these experiments suggest that it is
feasible to base more sophisticated formalizations of graph theory on algebraic
graphs. We did not experiment with adding additional axioms as described
in~\autoref{sec:algebra}. We do give an instantiation of the locale with the
unit type, illustrating that the current axioms are not sufficient.

@@

\section{Deep Embedding}\label{sec:deep}
Up until this point we were mainly concerned with exploiting the algebraic
structure of graphs for polymorphic graph constructions and transformations, and
for formal verification. Deeply embedding the construction primitives in a
datatype is interesting in itself. The definition of the datatype is
straightforward:
\begin{minted}{haskell}
data Graph a = Empty
             | Vertex a
             | Overlay (Graph a) (Graph a)
             | Connect (Graph a) (Graph a)
\end{minted}
An instance for the type class can be defined easily, however, we cannot use the
derived \hs{Eq} instance for the datatype, as it would violate the graph axioms:
\hs{Overlay Empty Empty} and \hs{Empty} both represent $\eps$ but are not deemed
equal.
One solution is to fold the deep embedding into a concrete graph representation to
obtain a law-abiding \hs{Eq} instance:
\begin{minted}{haskell}
fold :: Graph g => Graph (Vertex g) -> g
fold Empty         = empty
fold (Vertex  x  ) = vertex x
fold (Overlay x y) = overlay (fold x) (fold y)
fold (Connect x y) = connect (fold x) (fold y)

@@

The quotient type also allows us to formally prove that it satisfies the axioms,
by instantiating the locale corresponding to the core type class. It also paves
the way for further formalization efforts on the deep embedding. However, this
also requires more work to become truly interesting, i.e.\ the discovery of algorithms that
work directly on the deep embedding or other advanced usages.

@@

Algebraic graphs as presented have been successfully used in practice, e.g.\ for
asynchronous circuit design~\cite{beaumont2017high} and acceleration of graph
processing using FPGAs~\cite{mokhov2019language}. In these examples
the available transformations are sufficient. An important direction for future
research is going beyond those (relatively simple) transformations. That means
to find and formulate algorithms either inside the core type class or directly
on the deep embedding. Berghammer et. al~\cite{berghammer2020relational}, as well
as, Guttmann and Robinson-O'Brien~\cite{Relational_Minimum_Spanning_Trees-AFP}
successfully employ algebraic approaches to formally verify graph algorithms.
However, they are using an algebra of relations as opposed to the one here
defined on graph construction primitives. Dolan~\cite{dolan2013fun} uses a
semiring on adjacency matrices to find paths in graphs.
Mokhov himself postulates that the algebraic approach allows to formulate graph
algorithms as systems of equations with unknowns. This may also be an approach
for finding novel graph algorithms.
